Title Text

### 2013-07-25 schemeworkshop13.tex
Draft bivariant arrays

\section{Introduction}

Typed Clojure is a gradual type system for Clojure, based on the
static portion of Typed Racket.
This paper gives an overview of Typed Clojure, concentrating on extensions
that could interest Typed Racket users.

@@

@@

>>>>

\section{Typed Clojure}

Typed Clojure is a gradual type system for Clojure, based on the
static portion of Typed Racket.

@@

\section{Higher-rank polymorphism}

Typed Clojure supports f-bounded polymorphism and higher-rank
types.

@@

\section{Java interoperability}

As Clojure supports Java interoperability, Typed Clojure has some
integration with Java's type system. Calls to Java methods from
Clojure can be type checked by Typed Clojure.

We currently do not support Java generics

\subsection{Arrays}

Supporting statically sound interactions with Java arrays is a goal
of Typed Clojure. This is complicated by Java's decision to make
arrays covariant in their argument, a well documented source of static
unsoundness.

Our basic approach is to make our array types \emph{bivariant}. Array types
look like \lstinline|(Array2 w r)| and
are reminicent of functions or pipes: having a contravariant parameter for input (writing)
and a covariant parameter for output (reading).

Most commonly, an array type is invariant in its parameter; it can
write and read input of the same type.
We can get the same effect by setting our input and output
parameters to the same type. For example, \lstinline|(Array2 Number Number)|
(or equivalently, \lstinline|(Array Number)|)
in Typed Clojure is similar to invariant array types of \lstinline|Number| in languages like Scala.

The biggest gain in using a separate input parameter is the ability
to specify \emph{read-only} arrays. Crucially, our type system features an
explicit bottom type \lstinline|Nothing|, allowing a read-only \lstinline|Number| array
to be of type \lstinline|(Array2 Nothing Number)|.

To realise why defining read-only arrays are useful, we need to examine
what makes array covariance unsound in Java.
Array covariance lets us lie about the type of an array so the consumer
of an array cannot tell the actual type of the array when examining a type
signature.

\begin{verbatim}
...
public static Number[] getNumberArray() {
  Number[] n = new Integer[10];
  return n;
}
...
\end{verbatim}

To the casual consumer \emph{getNumberArray} returns an array that can both
read and write \lstinline|Number|s. However it is clear from the implementation
that attempting to write say a \lstinline|Double| to this array will result
in a runtime error.

\begin{verbatim}
...
Number[] myArray = getNumberArray();
myArray[0] = 1.1;
/* Exception in thread "main" java.lang.ArrayStoreException: java.lang.Double */
...
\end{verbatim}

Notice that this is a runtime error, and Java's type system has not helped
statically prevent it.
This could cause a similar issue for other statically-typed languages offering
interoperability with Java.

To prevent these sorts of runtime exceptions in Typed Clojure, we declare
all arrays from unknown sources to be \emph{read-only}. Put differently,
the only way to define a writeable array is to create it in type-checked Clojure
code.

\begin{lstlisting}
(let [n (CovariantArray/getNumberArray)]
  (aset n 0 1.1))

; Polymorphic static method clojure.lang.RT/aset could not be 
; applied to arguments:
; Domains: 
;         (Array2 i o) clojure.core.typed/AnyInteger i
; 
; Arguments:
;         (Array2 Nothing java.lang.Number) int (Value 1.1)
; 
; with expected type:
;         Any
\end{lstlisting}

The type inferred for the local \lstinline|n| is (Array2 Nothing Number)
which tells the type system: it is never safe to write to this array, but
it is safe to assume \lstinline|Number|s can be read from this array.

To emphasise, this is a static type error. Errors like this help Clojure programmers
use foreign Java libraries more correctly.

\begin{verbatim}
Note that Java libraries are often large and complex and programmers will probably
enjoy the extra help from the type system.
\end{verbatim}
